---
title: "p8105_hw3_zx2527"
author: "Zihan Xiong"
date: "2025-10-11"
output: github_document
---

# Library load

```{r}
library(tidyverse)
library(janitor)
library(p8105.datasets)
library(lubridate)
library(patchwork)
library(scales)
```


# Problem 1

```{r}
data("instacart")
glimpse(instacart)
```
**Description of the dataset**

There are `r nrow(instacart)` rows and `r ncol(instacart)` columns,
and each of them represents a single product added to a user’s grocery order on the Instacart. 

Key variables include:

1. order_id (unique order to help identify)
2. product_id 
3. add_to_cart_order (order in which each product was added to cart)
4. reordered (the product has been repurchased by the same customer or not)
5. user_id   
6. eval_set (which evaluation set this order belongs in)
7. order_number 
8. order_dow  (the day of the week on which the order was placed)
9. order_hour_of_day (the hour of the day on which the order was placed)
10. days_since_prior_order (days since the last order, capped at 30, NA if order_number=1)
11. product_name 
12. aisle_id (aisle identifier)
13. department_id   
14. aisle 
15. department 

Description:

For example, in order #1, user 112108 purchased six items — including products with IDs 49302 and 11109 — as part of their fourth recorded order (order_number = 4) on a Thursday (order_dow = 4).

```{r}
aisle_counts = instacart |>
  count(aisle) |>
  arrange(desc(n))


n_aisles <- nrow(aisle_counts)
n_aisles

head(aisle_counts, 10)
```
Description:

There are **134** aisles and **fresh vegetables (150609 orders) and fresh fruits (150473 orders)** are the most items ordered from.

**Make a plot on aisles with more than 10000 items**
```{r}
instacart |>
  count(aisle) |>
  filter(n > 10000) |>
  mutate(aisle=fct_reorder(aisle,n)) |>
  ggplot(aes(x = aisle, y=n)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Aisle",
    y = "Number of items ordered",
    title = "Items ordered by aisle (>10,000 orders)"
  ) +
  theme_minimal()
```

```{r}
top3_tbl = instacart |>
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle, product_name) |>
  summarize(n = n(), .groups = "drop_last") |>
  arrange(aisle, desc(n)) |>
  mutate(rank_in_aisle = min_rank(desc(n))) |>
  filter(rank_in_aisle <= 3) |>
  arrange(aisle, rank_in_aisle) |>
  select(aisle, product_name, n)

knitr::kable(top3_tbl, digits = 0, col.names = c("Aisle", "Product", "Orders"))
```

```{r}
mean_hour_tbl = instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(
    names_from = order_dow,
    values_from = mean_hour
  )

knitr::kable(mean_hour_tbl, digits = 2,
             col.names = c("Product", "Sun","Mon","Tue","Wed","Thu","Fri","Sat"))
```

# Problem 2

## data import
```{r}
zori_raw = 
  read_csv("./data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv",
           na = c("NA", "", "."), show_col_types = FALSE) |>
  janitor::clean_names()

zipcodes_raw = 
  read_csv("./data/Zip Codes.csv", na = c("NA", "", "."), show_col_types = FALSE) |>
  janitor::clean_names()
```

```{r}
names(zori_raw)
names(zipcodes_raw)
```

## clean and standardize zipcodes_raw
```{r}
zipcodes_raw = 
  zipcodes_raw |>
  mutate(
    zip = str_pad(as.character(zip_code), 5, pad = "0"),
    county = str_to_title(county) |> str_trim()
  )

county_to_borough = c(
  "Bronx"      = "Bronx",
  "Kings"      = "Brooklyn",
  "New York"   = "Manhattan",
  "Queens"     = "Queens",
  "Richmond"   = "Staten Island"
)

zipcodes_raw = 
  zipcodes_raw |>
  mutate(borough = recode(county, !!!county_to_borough)) |>
  filter(!is.na(borough)) |>
  distinct(zip, .keep_all = TRUE) |>
  select(zip, borough, county, neighborhood)
```

## pivot longer for zori_raw
```{r}
zori_raw = 
  zori_raw |>
  mutate(zip = str_pad(as.character(region_name), 5, pad = "0")) |>
  pivot_longer(
    cols = matches("^x\\d{4}_\\d{2}_\\d{2}$"),
    names_to = "date",
    values_to = "rent"
  ) |>
  mutate(
    date = ymd(str_replace_all(str_remove(date, "^x"), "_", "-")),
  ) |>
  filter(!is.na(date))

zori_raw
```

## join the tables
```{r}
zillow_nyc = 
  left_join(zori_raw, zipcodes_raw, by = "zip") |>
  filter(!is.na(borough)) |>
  select(zip, borough, county, neighborhood, date, rent)

zillow_nyc

```

## restrict to Jan 2015–Aug 2024 
```{r}
zillow_2015_2024 =
  zillow_nyc |>
  filter(date >= as.Date("2015-01-01"),
         date <= as.Date("2024-08-31"))

zip_coverage =
  zillow_2015_2024 |>
  group_by(zip) |>
  summarize(n_months = n_distinct(date), .groups = "drop")

n_zip_116  = zip_coverage |> filter(n_months == 116) |> nrow()
n_zip_lt10 = zip_coverage |> filter(n_months < 10)   |> nrow()

n_zip_116
n_zip_lt10
```
There are 149 ZIP codes are observed 116 times, and 0 of them are observed fewer than 10 times.

ZIPs observed rarely are typically edge cases (boundary changes, small or newly defined ZIPs, or missing data), whereas ZIPs observed in every month are stable residential/commercial areas with consistent reporting.

## Derive borough from county and make the borough × year table
Description:

This table summarizes the average Zillow Observed Rent Index (ZORI) across New York City boroughs for 2015–2024. Values are rounded to the nearest dollar for readability.

Boroughs such as Manhattan and Brooklyn show consistently higher mean rent compared to Bronx or Staten Island, and all boroughs display an upward trend in recent years (especially post-2021).

```{r}
borough_year_tbl =
  zillow_2015_2024 |>
  mutate(year = lubridate::year(date)) |>
  group_by(borough, year) |>
  summarize(
    mean_rent = mean(rent, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(borough, year)

borough_year_wide =
  borough_year_tbl |>
  pivot_wider(
    names_from = year,
    values_from = mean_rent
  )|>
  mutate(across(-borough, ~ round(.x, 0))) 

borough_year_wide |>
  knitr::kable(
    digits = 0,
    caption = "Average Monthly Rent by Borough and Year (2015–2024)"
  )

```

## Plot of NYC Rental Prices across ZIP codes and Boroughs

Description:

This plot visualizes monthly rent trends (ZORI) for all NYC ZIP codes between 2015–2024. Thin lines show individual ZIP code trajectories; bold lines indicate borough-level medians.

In general, Manhattan and Brooklyn have consistently higher rents with strong growth post-2021, and all of the boroughs showed an increased trend.
```{r}

zillow_clean = 
  zillow_2015_2024 |> 
  filter(!is.na(rent), is.finite(rent)) |> 
  group_by(borough, date) |> 
  mutate(median_rent_borough = median(rent, na.rm = TRUE)) |> 
  ungroup()

all_years_plot = zillow_clean |> 
  ggplot(aes(x = date, y = rent, group = zip)) +
  geom_line(alpha = 0.1) +
  stat_summary(
    aes(group = borough),
    fun = median,
    geom = "line",
    linewidth = 1
  ) +
  facet_wrap(~ borough, ncol = 2, scales = "free_y") +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "NYC Rental Prices by Borough, 2015–2024",
    subtitle = "Thin lines = ZIP codes; Thick lines = Borough medians",
    x = "Year",
    y = "ZORI (Rent, USD)"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

all_years_plot
```

## Violin plot for rent distribution by borough
```{r}

rent_2023_zip = 
  zillow_2015_2024 |> 
  mutate(year = year(date)) |> 
  filter(year == 2023) |> 
  group_by(borough, zip) |> 
  filter(!is.na(rent), is.finite(rent))|>
  summarize(
    mean_rent = mean(rent, na.rm = TRUE),
    .groups = "drop"
  )

rent_2023_plot = 
  rent_2023_zip |> 
  ggplot(aes(x = borough, y = mean_rent)) +
  geom_violin(alpha = 0.6) +
  labs(
    title = "Distribution of Average ZIP-level Rents by Borough (2023)",
    x = "Borough",
    y = "Average Monthly Rent (USD)"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

rent_2023_plot
```

## combine
```{r}
combined_plot =
  (all_years_plot | rent_2023_plot) +
  patchwork::plot_annotation(
    title = "NYC Rental Prices: Long-term Trends and 2023 Distributions",
    subtitle = "Left: Borough-level rent trends (2015–2024)  |  Right: ZIP-level average rent distributions (2023)"
  )


combined_plot


dir.create("results", showWarnings = FALSE)
ggsave(
  filename = "results/combined_rent_panels.png",
  plot = combined_plot,
  width = 13, height = 6, dpi = 300
)


```

## Problem 3
```{r}

```